Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
two questinos about the code
filled two 'holes' and added to the comparison section

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
We~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

@@

We hope you find the search code simple and straightforward:
\smallfuzzverbatiminput{11pt}{search}
\emph{[MENTION AND QUANTIFY NOAH'S SUCCESSES IN ADDING NEW HEURISTICS.]}

@@

Our other ugly code came from  attempts to debug
run-time errors.
MRFy uses a Markov random field which is represented in a legacy file
format that uses a poorly documented representation---when beta
strands overlap and are doubly paired, the invariants of the
representation are not clear.
Dealing with this representation is difficult no matter what
programming language is used, and faults are to be expected.
Using GHC, we had a hard time diagnosing run-time errors.
Extensive use of 
\texttt{trace} 
littered our code,
even when moved from our main code to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, the results weren't good enough.
\emph{[WHY WASN'T THE BACKTRACE FEATURE GOOD ENOUGH?]}
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

@@

When performance has mattered, the members of our group, like other
computational biologists, have used~C++.
Our~prior experience therefore includes building,
maintaining, and extending research tools written in~C++.
One~such tool is~Matt,
a~structural protein aligner~\cite{Menke:2008wu}.
Matt comprises about 12,000 lines of~C++.
It uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.

We~modified Matt to use information about sequences as well as structure.
We~thought the modification would take three months, 
but it took most of a~year.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: 
nearly every change resulted in new segfaults.
We~had hoped to add support for partial alignments,
which we~expected to require only 
a cosmetic manipulation of the 
output, but
this feature wound up
 requiring deep information about some data structures,
and we had to give~up.
We~believe we could write an equivalent tool in~Haskell,
with most of Matt's performance, in at most nine months.

Like many researchers, we~and our colleagues often write code hastily.
Hastily written C++ code can get a job done, 
but we have found it expensive in the long run.
Our most painful experience was with an implementation of
\textit{multidimensional}
dynamic programming which was hastily added to the SMURF homology detector
\cite{Menke:2010ti}.
The implementation took 12,000~lines, including a single function of
over 700 lines.
When we added simulated evolution to this code, a relatively minor
enhancement \cite{Daniels:2012}, 
just understanding the existing code took several months.
For comparison, 
building the first functioning version of MRFy took three months of part-time work.
And we expect that higher-order functions will make extending MRFy
easy.
\remark{I'm thinking this \P\ is not so strong. ---NR}

>>>>

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
We~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
\remark{Two questions: (1)~Why isn't the function called \texttt{vee}?
(2)~Do we prefer \texttt{minimum (map from [Mat, Ins, Del])}?
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

@@

We hope you find the search code simple and straightforward:
\smallfuzzverbatiminput{11pt}{search}

We found it easy to write several stochastic search heuristics thanks to
this framework.
We first implemented random-mutation hill climbing in one day and roughly
fifty lines of code.
We next implemented simulated annealing, and thanks to Haskell's module system,
we imported the \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} functions
from the random-mutation hill climbing module.
Indeed, we had only to write a new \texttt{accept} function, comprising only ten
lines of code and half an hour of our time.
We then implemented a genetic algorithm approach; this was still able to import
existing \texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
function was only a few lines of code.
Understandably, though, the complexities of representing ``crossover'' of parent
placements required forty lines of code and much of a day of thought.

@@

Our other ugly code came from  attempts to debug
run-time errors.
MRFy uses a Markov random field which is represented in a legacy file
format that uses a poorly documented representation---when beta
strands overlap and are doubly paired, the invariants of the
representation are not clear.
Dealing with this representation is difficult no matter what
programming language is used, and faults are to be expected.
Using GHC, we had a hard time diagnosing run-time errors.
Extensive use of 
\texttt{trace} 
littered our code,
even when moved from our main code to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, the results weren't good enough.
Specifically, because the backtrace feature required profiling mode,
it could only be used in very small test cases that didn't always trigger
run-time errors, and one of us was unable to overcome difficulties in building
various libraries in profiling mode.
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

@@

When performance has mattered, the members of our group, like other
computational biologists, have used~C++.
Our~prior experience therefore includes building,
maintaining, and extending research tools written in~C++.
One~such tool is~Matt,
a~structural protein aligner~\cite{Menke:2008wu}.
Matt comprises about 12,000 lines of~C++.
It uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
It relies on no external libraries beyond \texttt{zlib},
but contains 1,000 lines of protein-structure code that ought
to be in a library.

We~modified Matt to use information about sequences as well as structure.
We~thought the modification would take three months, 
but it took most of a~year.
This modification added 2,000 lines of code, and relies on
system calls to external programs (sequence aligners) that we did not write.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: 
nearly every change resulted in new segfaults.
We~had hoped to add support for partial alignments,
which we~expected to require only 
a cosmetic manipulation of the 
output, but
this feature wound up
 requiring deep information about some data structures,
and we had to give~up.
We~believe we could write an equivalent tool in~Haskell,
with most of Matt's performance, in at most nine months.

Like many researchers, we~and our colleagues often write code hastily.
Hastily written C++ code can get a job done, 
but we have found it expensive in the long run.
Our most painful experience was with an implementation of
\textit{multidimensional}
dynamic programming which was hastily added to the SMURF homology detector
\cite{Menke:2010ti}.
The implementation took 9,000~lines, including a single function of
over 700 lines.
When we added simulated evolution to this code, a relatively minor
enhancement \cite{Daniels:2012}, 
just understanding the existing code took several months.
Part of SMURF's code includes the 1,000 lines of protein-structure code
copied from the previous Matt software.

For comparison, 
building the first functioning version of MRFy took three months of part-time work.
And we expect that higher-order functions will make extending MRFy
easy.
MRFy currently contains 1,200 lines of Haskell code, though it relies heavily on
a number of libraries beyond the standard Prelude.
We found that adapting our software to explore a new search heuristic was typically
less than a day's work.
There are two major research ideas we wish to add to MRFy eventually, one of which
\remark{how much should I say about profiles and alpha helices? ---NMD}
involves looking at another type of nonlocal dependency (\textit{alpha helices} in contrast
to the current beta strands).
We expect that thanks to algebraic data types and typeclasses, these enhancements will be at
most a few weeks of work; we do not believe these enhancements would even be feasible in SMURF.
Indeed, Daniels tried to add another of these enhancements to SMURF two years ago and gave up due
to the difficulty of the implementation.
\remark{The above needs a lot of revision. It is a shitty first draft of a /P/. ---NMD}
